---
title: "Introduction to Incomplete Matrix Regression (IMR)"
output: rmarkdown::html_vignette
description: >
  Learn how to use IMR package for matrix regression. We will go over how to create the data object, select the model structure,
  how to choose the model's hyperparameters, and interpret the output and assess the performance of the model.
vignette: >
  %\VignetteIndexEntry{Introduction to IMR}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
#library(IMR)
library(dplyr)
library(magrittr)
set.seed(2026)
#source("../article/bixi/data_bixi.R")
#source("../article/bixi/helpers_bixi.R")
```

IMR is feasible on any data that can be expressed in the form of a matrix of real numbers. Moreover, there is a wide range of accepted side information. We will begin by discussing the data and how to prepare the data for our model.

## Data

As an example, we will simulated data from the distributions as our Simulation Studies section in the paper. We will simulated a matrix of dimension $500\times 600$ of rank 5 and with $80\%$ of it being missing. We will simulate 3 row covariates that change from on row to another of dimension $500\times 3$. We will simulated 4 column covariates that are shared among the columns, with dimension $4\times 1$.


As our example, we will use the BIXI [^1], a docked bike-sharing service in Montreal, Canada. We use the data compiled by [@lei2025], which contains daily departure counts for each of 587 stations over 196 days (April 15 to October 27, 2019). The data is normalized between 0 and 1. The data is accompanied with side information. We will first

[^1]: website: <https://bixi.com/>

```{r echo=FALSE, eval=FALSE}
#=================
dat <- prepare_bixi_data()
X <- dat$X %>% as.matrix()
Z <- dat$Z %>% as.matrix()
Y <- dat$modd$Y
kernels <- generate_similarity_bixi(0.25, "Feb_last", prefix = "1",
                                      train_prefix = "65",
                                    "./article/bixi/data/splits2")
bixi_example <- list(Y=Y, X=X,Z=Z, kernels = kernels, test = dat$test)
usethis::use_data(bixi_example)

saveRDS(data_obj, "./data/bixi_example.rda")
saveRDS(dat$test, "./data/bixi_test.rda")
#----------------------------------------------
```


```{r}
bixi <- IMR::bixi_example

head(bixi$X)
head(bixi$Z)
dim(bixi$Y)
bixi$Y[1:10,11:12]

print(length(bixi$test@x)/length(bixi$test))
print(length(bixi$Y@x)/length(bixi$Y))
```


```{r}

data <- IMR::imr_data(bixi$Y, bixi$X, bixi$Z, val_prop = 0.2, seed = 2026)

data

```

## Model Structure

```{r}
data <- update(data, shared_gamma = TRUE, col_intercept = TRUE, row_intercept = TRUE)
print(data)

```


## Fit function


```{r}
convergence <- IMR::imr_convergence(maxit = 600, thresh=1e-4, trace = TRUE)
fit <- IMR::imr_fit(data, rank = 10, lambda_m = 1e-1, lambda_beta = .02, lambda_gamma = 0)
print(fit)
summary(fit)
names(coef(fit))

```



## Interpretting the results and assessing the performance

```{r}

data_out <- IMR::reconstruct(fit, data)

print(names(data_out))


data_out$xbeta[1,1] + data_out$gammaz[1,1] + data_out$M[1,1] + coef(fit)$beta0[1] +
  coef(fit)$gamma0[1]

```


```{r}
preds <- IMR::reconstruct_partial(fit, data, bixi$test, trace=TRUE)
print(preds[1:10,1:10])
print("Test Error")
print(IMR::evaluate(preds@x, bixi$test@x))
```


```{r}
preds <- IMR::reconstruct_partial(fit, data, data$y_train, trace=TRUE)
print("Training Error")
print(IMR::evaluate(preds@x, data$y_train@x))

```



## Hyperparameter Tuning

```{r}
grid <- IMR::imr_tune_grid(
  beta = c(0, NA, 40), # (min, max, length)
  gamma = c(0, NA, 40), #(min, max, length)
  laplace = c(0, NA, 40, 2), #(min, max, length, early-stopping tolerance)
  rank = c(2, 15, 1) #(min, max, step)
)
print(grid)

```


```{r}
grid <- IMR::imr_set_grid_limits(data, grid, verbose = TRUE)
print(grid)

```

```{r}
cv_out <- IMR::imr_tune(data, grid, n_cores=5, seed = 2026, verbose=1)
summary(cv_out$fit)
print(cv_out$fit)
```


```{r}
preds <- IMR::reconstruct_partial(cv_out$fit, data, bixi$test, trace=TRUE)
print(IMR::evaluate(preds@x, bixi$test@x))

  
```



## Similarity Matrices and within-rows and within-columns autocorrelation

```{r}
similarity_cols <- IMR::imr_similarity(bixi$kernels$spatial, invert=FALSE, jitter = 0.01)
similarity_rows <- IMR::imr_similarity(bixi$kernels$temporal, invert=FALSE, jitter = FALSE)
print(similarity_rows)
print(similarity_cols)

#IMR::imr_similarity("matern", d = ..., matern_params = c(smoothness= 2.5, range = 1))
```


```{r}
data <- IMR::imr_data(bixi$Y, bixi$X, bixi$Z, 
                      similarity_rows = similarity_rows,
                      similarity_cols = similarity_cols,
                      val_prop = 0.2, seed = 2026)

data <- update(data, shared_gamma = TRUE, col_intercept = TRUE, row_intercept = TRUE)
print(data)

grid <- IMR::imr_tune_grid(
  beta = cv_out$params$lambda_beta,
  gamma = cv_out$params$lambda_gamma,
  laplace = c(0, NA, 40, 2), #(min, max, length, early-stopping tolerance)
  rank = c(2, 15, 1) #(min, max, step)
)
print(grid)
grid <- IMR::imr_set_grid_limits(data, grid, verbose = TRUE)
cv2_out <- IMR::imr_tune(data, grid, n_cores=9, seed = 2026, verbose=1)
summary(cv2_out$fit)
print(cv2_out$fit)
preds <- IMR::reconstruct_partial(cv2_out$fit, data, bixi$test, trace=TRUE)
print(IMR::evaluate(preds@x, bixi$test@x))


```

