---
title: "Introduction to Incomplete Matrix Regression (IMR)"
output: rmarkdown::html_vignette
description: >
  Learn how to use IMR package for matrix regression. We will go over how to create the data object, select the model structure,
  how to choose the model's hyperparameters, and interpret the output and assess the performance of the model.
vignette: >
  %\VignetteIndexEntry{Introduction to Incomplete Matrix Regression (IMR)}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
---

```{r, echo = FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(IMR)
library(dplyr)
library(magrittr)
set.seed(2026)

```

The IMR package provides a framework for matrix regression on incomplete matrices. Let $\boldsymbol{Y} \in \Re^{n\times m}$ denote the observed incomplete matrix, where missing values are designated by either `NA` or zero. The estimator for each entry of the matrix is specified by the model

$$
\boldsymbol{\hat Y_{ij}} = \boldsymbol{\hat\beta_{oi}} + \boldsymbol{\hat\Gamma_{oj}} + \boldsymbol{X_i\hat\beta_j} + \boldsymbol{\hat\Gamma_i Z^{'}_j} + \boldsymbol{\hat M_{ij}}
$$

where $\boldsymbol{X}\in \Re^{n\times p} and  \boldsymbol{Z}\in \Re^{m\times q}$ are matrices of $p$ row and $q$ column covariates, respectively. The vectors$\boldsymbol{\hat\beta_{o}}\in \Re^{n}$ and $\boldsymbol{\hat\Gamma_{o}}\in \Re^{n}$ represent the row-level and column-level intercepts. The term $\boldsymbol{\hat\beta}$ denotes the row covariate coefficients, which may be structured as either an $p\times m$ matrix, allowing coefficients to vary across columns, or a $p$-dimensional vector, forcing coefficients to be equal across all columns for each covariate. Similarly, the column covariate coefficients, denoted by $\boldsymbol{\hat\Gamma}$, can be either an $n \times q$ matrix or a $q$-dimensional vector where all rows share the same coefficient for each covariate. Finally, $\boldsymbol{M}$ is a rank-r (r is a hyperparameter) matrix subject to a nuclear norm penalty. Lasso (\$L_1\$) penalties are imposed on the row and column covariates as well, yielding the following penalty structure:

$$
\mathrm{Penalty} = \lambda_\beta {\|\boldsymbol{\beta}\|}_{1} + \lambda_\Gamma {\|\boldsymbol{\Gamma}\|}_{1} + \lambda_m {\|\boldsymbol{S_r^{1/2}MS_c^{1/2}}\|}_{*},
$$

where $\boldsymbol{S_r} \in \Re^{n\times n}$ and $\boldsymbol{S_c} \in \Re^{m \times m}$ are similarity (or information) matrices that describe the correlation structure among the rows and columns of the response matrix, respectively. In the absence of a known correlation structure, this penalty term reduces to the standard nuclear norm, ${\|\boldsymbol{M}\|}_{*}$.

Because it is not always necessary or feasible to incorporate all components simultaneously, the framework permits the estimation of any designated sub-model. Examples of permissible specifications include:

1.  $\boldsymbol{\hat Y_{ij}} = \boldsymbol{\hat\beta_{oi}} + \boldsymbol{X_i\hat\beta}$ (where $\boldsymbol{\hat\beta}$ is a p-dimensional vector).
2.  $\boldsymbol{\hat Y_{ij}} = \boldsymbol{\hat\beta_{oi}} + \boldsymbol{\hat\Gamma_{oj}} +\boldsymbol{\hat M_{ij}}$ with $\boldsymbol{S_r}$ set to a the inverse of a Matern kernel and $S_c$ left unspecified.
3.  $\boldsymbol{\hat Y_{ij}} = \boldsymbol{\hat M_{ij}}$, where neither $\boldsymbol{S_r}$ or $S_c$ is specified. This corresponds exactly to the Soft-Impute model [@hastie2015].

This vignette is structured as follows: we first discuss the data structure and pre-processing steps for the proposed model, using an illustrative example. Second, we show the specification of the model architecture and how we modify it. Third, we demonstrate the application of the fitting function under known regularization parameters, followed by a presentation of the results and performance assessment. Next, we describe our procedure for regularization parameters' tuning. We conclude with a discussion on the construction and integration of the similarity matrices, $\boldsymbol{S_r}$ and $\boldsymbol{S_c}$.

## Data

As an example, we analyze data from Bixi [^1], a docked bike-sharing service in Montreal, Canada. We use the data compiled by [@lei2025], which contains normalized daily departure counts for each of 587 stations over 196 days (April 15 to October 27, 2019). The data is accompanied by side information. We select two temporal variables (temperature and precipitation) and one spatial variable (park area). Below is a summary of the data (Note that the dataset also contains similarity matrices which we we discuss them in the final section).

[^1]: website: <https://bixi.com/>

```{r}
bixi <- IMR::bixi_example

head(bixi$X,2)
head(bixi$Z,3)
dim(bixi$Y)
bixi$Y[1:6,11:12]

sprintf("Percentage of observed entries in Y: %.1f%%", 100*sum(bixi$Y != 0)/length(bixi$Y))
sprintf("Percentage of entries in the test set: %.1f%%", 100* sum(bixi$test != 0)/length(bixi$test))
```

To facilitate data integration within the proposed framework, we introduce the `imr_data` class, which automatically structures the inputs into a model-compliant format. As in the preceding output, unobserved entries in the response matrix are denoted by $0$, though `NA` is also supported. Because we treat all zero-values entries as missing data, structural zeros cannot be accommodated as observed values. 

```{r}

data <- imr_data(Y = bixi$Y, X =  bixi$X, Z =  bixi$Z, val_prop = 0.2, seed = 2026)
print(data)

```

The response argument `Y` in the function above accepts any standard matrix class from base R or the `Matrix` package. The framework coerces this input into `Incomplete` class, a subclass of `dgCMatrix`, motivated by the analogous class in the `SoftImpute` package. Any matrix can be cast to this class as follows:
```{r}
bixi$test <- as.Incomplete(bixi$test)
bixi$test[3:5,3:4]
```
Furthermore, the function performs a QR decomposition on both covariate matrices, $\boldsymbol{X}$ and $\boldsymbol{Z}$, storing the resultant components as `XQ`, `XR`, `ZQ` and `ZR`. This transformation is necessary as the fit function works on orthonormal matrices (see the main article for more details). However, in our final coefficient estimates, the estimates are projected back onto the original covariate spaces.

## Model Structure

The latter portion of the preceding data summary shows the model configuration. The first column contains the supplied auxiliary information, indicating, for instance, the presence of two row convariates and one column covariate. The second column specifies the active components in the estimator. By default, the low-rank matrix and any provided side information are included in the active model specification. The designation `(Unshared)` indicates that the covariate coefficients are permitted to vary across rows or columns; that is, the coefficients take the form of matrices rather than vectors. This default specification can be modified via the `update()` function. To illustrate, we introduce row intercepts and constrain the row covariate coefficients to be shared across all columns, therby enforcing a fixed effect of temperature and precipitation across all stations.

```{r}
data <- update(data, shared_beta = TRUE, row_intercept = TRUE)
print(data)
```

The `update` function modifies the model architecture through logical arguments that activate or deactivate specific components. We proceed to fit this model.

## Fit function

The fit function requires fixed values for the regularization parameters: the three penalty terms and the rank of the low-rank component. For now, let's set artibrary values for these parameters. Tuning procedure is detailed ina  subsequent section.

```{r}
fit <- imr_fit(data, rank = 10, lambda_m = 1e-1, lambda_beta = 0, lambda_gamma = 0.02)
```

The `print` function yields diagnostic information regarding the estimation procedure, while the `summary` function provides an overview of the resultant estimates. The estimated coefficients can be extracted using `coef` function or directly by accessing the `fit$coefficients` object. It should be noted that the `summary` function show the covariate coefficients with respect to the original covariate matrices $\boldsymbol{X}$ and $\boldsymbol{Z}$, while the raw extracted coefficients in `coef()` correspond to the orthonormal matrices `XQ` and `ZQ`.

```{r}
print(fit)
summary(fit)
names(coef(fit))
```


Convergence criteria, specifically the maximum allowed iterations and the tolerance threshold are specified using the `imr_convergence` object:

```{r}
convergence <- imr_convergence(maxit = 15, thresh = 1e-5, trace = TRUE); print(convergence)
fit2 <- imr_fit(data, rank = 10, lambda_m = 1e-1, lambda_beta = 0, lambda_gamma = 0.02,
                convergence = convergence)
```

In the output above, `obj` denotes the value of the objective function at the current iteration and `ratio` represents the relative change in the parameter estimates between successive iterations, evaluated using the Frobenius norm. The optimization algorithm terminates when either the maximal iteration limit is attained or the computed ratio falls below the pre-specified tolerance threshold.

## Interpreting Results and Assessing Performance

We begin by showing the two imputation methods we have, beginning with the full reconstruction of the response matrix.

```{r}
data_out <- reconstruct(fit, data)
print(names(data_out))
```

The `reconstruct` function outputs all estimated model components alongside the fully imputed response matrix (`$estimates`). It should be noted that $\beta$ and $\Gamma$ estimates are projected back onto the original covariate matrices, $X$ and $Z$. For illustration, the explicit computation for the first element of the response matrix is

```{r}
cat(sprintf("True value of the entry (1,1) int he data matrix is %.4f and the estimated value is %.4f",
        data$Y[1,1],
        data_out$beta0[1] + # row intercept
        data_out$gamma0[1] + # column intercept
        bixi$X[1,] %*% data_out$beta + # row covariates
        data_out$gamma[1,] %*% t(bixi$Z)[,1] + # column covariates
        data_out$M[1,1]))# low-rank matrix

```

This reconstruction function can be computationally prohibitive in high-dimensional settings. Moreover, we may only be interested in imputing a subset of the response matrix. Consider the test dataset we have. Let's only impute the entries that corresponds to that set.

```{r}
preds <- reconstruct_partial(fit, data, bixi$test@i, bixi$test@p, trace=TRUE, return_matrix = TRUE)
print(preds[1:10,1:10])
```

The `reconstruct_partial` function takes two additional arguments, the $i$ and $p$ vectors, which denote the row and column indicies of the entries to be estimated. If the `return_matrix` argument is `TRUE`, the function returns a sparse matrix containing the estimates. If it's `FALSE`, then it returns a vector of the estimates corresponding directly to the `@x` slot of the sparse data structure. We also provide an evaluation function that compute standard error metrics between two numeric vectors.

```{r}
knitr::kable(evaluate(preds@x, bixi$test@x),format = "pipe",digits = 4)
```

## Hyperparameter Tuning

We now address the tuning of the regularization hyperparameters. Our model incorporates four parameters: three regularization weights $(\lambda)$ and the rank of the low-rank component, $r$. We begin by constructing a parameter grid. For the lambdas ($\lambda_\beta,\lambda_\Gamma, \lambda_m$), the grid is defined by a minimum, a maximum, and a sequence length. For the rank, the grid requires a minimum, a maximum, and a step size. Both the nuclear norm penalty and the rank require an early-stopping tolerance parameter; a value of 2 or 3 is generally sufficient. If the maximum for any of the lambdas is unknown, it may be initialized to `NA`. A dedicated function is subsequently used to estimate these upper bounds using the Karush-Kuhn-Tucker (KKT) conditions.


```{r}
grid <- imr_tune_grid(
  beta = c(0, NA, 40), # (min, max, length)
  gamma = c(0, NA, 40), #(min, max, length)
  nuclear = c(0, NA, 40, 2), #(min, max, length, early-stopping tolerance)
  rank = c(2, 15, 1,2) #(min, max, step, early-stopping tolerance)
)
print(grid)

```

To fix any parameter to a specific constant, it can be supplied as a scalar. Parameters omitted from the specification will automatically default to standard configurations, as illustrate below:
```{r}
imr_tune_grid(
  beta = 3, 
  gamma = c(0, NA),
  nuclear = 0,
  rank = c(0, 10)
)
```
Returning to the initial grid specification, we need to find upper limits for the $\lambda$ parameters. This is achieved with a two-step procedure for each unspecified maximum:

1- An initial upperbound is derived using the KKT conditions.
2- Because the KKT bound can occasionally be too big, we perform a subsequent bisection search to find the minimum $\lambda$ required to shrink all corresponding coefficients exactly to zero.

When determining the maximum for either $\lambda_\beta$ or $\lambda_\Gamma$, the penalty for the alternate covariate is set to infinity, while the low-rank hyperparameters are held at default values. Conversly, when determining the upper limit for $\lambda_m$, the two covariate penalties are fixed at default values.

```{r}
convergence <- imr_convergence(maxit = 1000, thresh = 1e-5)
grid <- imr_set_grid_limits(data, grid,
                            default_rank = 5, default_lambda_m = 0, 
                            default_lambda_beta = 0, default_lambda_gamma = 0,
                            bisection_iter = 10, # number of iteration in the second step
                            verbose = TRUE,convergence = convergence )

print(grid)
```
With the grid fully specified, we proceed to find the optimal hyperparameters. The tuning function internally runs one of the three estimation processes:

1. Tuning $\lambda_m$ and $r$ only: if both $\lambda_\beta$ and $\lambda_\Gamma$ do not require tuning, then the procedure optimizes the low-rank component parameters sequentially using one of the two following modes:

- Fast Mode: Initializes at the maximum $\lambda_m$ and minimum rank $r$ then simultaneously decreases $\lambda_m$ and increases $r$ at each iteration. We terminate when validation performance cease to improve, governed by the early-stopping tolerance of $\lambda_m$.
- Slow Mode: Constructs a fully nested grid. For every candidate value in $\lambda_m$, it sequentially evaluates ranks in ascending order until validation performance degrades, governed by the early-stopping tolerance of the rank.

2. One Covariate:  if only one of $\lambda_\beta$ and $\lambda_\Gamma$ requires tuning, the algorithm parallelizes across the respective covariate grid. For each candidate value, it executes the first scheme above to determine the optimal low-rank parameters.

3. Alternating Optimization: If both $\lambda_\beta$ and $\lambda_\Gamma$ require tuning, an alternating minimization procedure is employed. The algorithm fixes $\lambda_\Gamma$ and tunes $\lambda_\beta$ (along with $\lambda_m$ and $r$) using scheme 2 above; it then fixes $\lambda_\beta$ and tunes $\lambda_\Gamma$, again using scheme 2. It iterates until the difference in estimated coefficients is smaller than some threshold.

Given that all parameters in our example require tuning, the algorithm uses the third scheme. We use the fast mode, which generally yields good performance. For an exhaustive final tuning, however, the slow model is recommended, as it evaluates the complete the parameter space of $\lambda_m$ and $r$.

```{r}
cv_out <- imr_tune(data, grid, fast_nuclear = TRUE, convergence = convergence,
                   n_cores=9, seed = 2026, verbose=1)
summary(cv_out$fit)
print(cv_out$fit)
```

```{r}
preds <- reconstruct_partial(cv_out$fit, data, bixi$test@i, bixi$test@p, return_matrix = FALSE)
knitr::kable(evaluate(preds, bixi$test@x),format = "simple",digits = 4)
```

## Similarity Matrices and within-Row and Within-Column Autocorrelation

We now turn to the final component of the proposed methodology: the incorporation of similarity matrices. In many applications, the response matrix exhibits correlation structures across its rows, its columns, or both. These structures can be explicitly described with similarity or covariance matrices. The framework adjusts the estimates of the low-rank component for these correlation structures.

Returning to the BIXI dataset, the data are supplemented by two auxiliary distance matrices: a spatial distance matrix derived from the geographic coordinates of the stations (columns) and a temporal distance matrix representing the elapsed days (rows). For the prpose of this illustration, we impose a Matern 5/2 covariance structure on the columns and a Gaussian (radial basis function) kernel with a length-scale of 6 on the rows. These kernels are defined using the `imr_similarity` function.

```{r}
similarity_cols <- imr_similarity(x = "matern", d = bixi$spatial_distance, invert=TRUE, jitter = .1,
                                  matern_smoothness = 5/2, matern_range = .018); print(similarity_cols)

similarity_rows <- imr_similarity(x = "rbf", d = bixi$temporal_positions, invert=TRUE, jitter = .1,
                                  rbf_ell = 6); print(similarity_rows)


```

Key notes regarding the `imr_similarity` function:

1- The primary argument, `x`, accepts either a character string ("Matern" or "rbf") or a precomputed covariance or information matrix. If a string is provided, a corresponding distance matrix `d` must be supplied. If a precomputed matrix is supplied, then `d` is omitted.

2- The framework uses an information (precision) matrix. Consequently, if a covariance kernel is specified or provided, the argument `invert=TRUE` must be declared to compute the inverse.

3- Setting `jitter` to small positive value help reducing the condition number. The supplied value will be added to the diagonal elements of the covariance matrix before inverting it.

4- The functions returns the components `U` and `d`, corresponding to the eigenvectors and eigenvalues of the target similarity matrix, respectively. This representation is the required input for the framework.

We proceed to integrate these matrices into the data object and update the model specification

```{r}
data <- imr_data(bixi$Y, bixi$X, bixi$Z, 
                      similarity_rows = similarity_rows,
                      similarity_cols = similarity_cols,
                      val_prop = 0.2, seed = 2026)

data <- update(data, shared_beta = TRUE, row_intercept = TRUE)
print(data)
```

We rerun the hyperparamter tuning. However, this time, we only tune $\lambda_m$ and $r$ as the values for $\lambda_\beta$ and $\lambda_\Gamma$ remain fixed to their previously determined optimal values.
```{r}
grid <- imr_tune_grid(
  beta = cv_out$params$lambda_beta,
  gamma = cv_out$params$lambda_gamma,
  nuclear = c(0, NA, 40, 2),
  rank = c(2, 15, 2) 
)
print(grid)
grid <- imr_set_grid_limits(data, grid, verbose = TRUE)
cv2_out <- imr_tune(data, grid, n_cores=9, seed = 2026, verbose=1)
summary(cv2_out$fit)
print(cv2_out$fit)
```

```{r}
preds <- reconstruct_partial(cv2_out$fit, data, bixi$test@i, bixi$test@p)
knitr::kable(evaluate(preds, bixi$test@x),format = "pipe",digits = 4)

```
